R-LIME: LIME法における矩形制約と最適化

### 2024-03-04 paper/paper.tex
ICPR 2nd version

\section{Proposed Method}
\begin{figure}[tbp]
	\centering
	\begin{subfigure}[t]{0.3\textwidth}
		\centering
		\includegraphics[width=\textwidth]{visual-lime}
		\caption{%
			LIME: Locally approximates the decision boundary around the focal point.
		}\label{fig:lime}
	\end{subfigure}%
	\hspace{0.03\textwidth}
	\begin{subfigure}[t]{0.3\textwidth}
		\centering
		\includegraphics[width=\textwidth]{visual-anchor}
		\caption{%
			Anchor: Maximizes coverage of a rectangular region
			containing the focal point under accuracy constraints.
		}\label{fig:anchor}
	\end{subfigure}
	\hspace{0.03\textwidth}
	\begin{subfigure}[t]{0.3\textwidth}
		\centering
		\includegraphics[width=\textwidth]{visual-rlime3}
		\caption{%
			R-LIME:
			Maximizes coverage of a rectangular region
			under lower constraints on the accuracy of the linear classifier.
		}\label{fig:rlime}
	\end{subfigure}
	\caption[Visual comparison of LIME, Anchor and R-LIME]{%
		Visual comparison of LIME, Anchor and R-LIME (our proposed method).
		The solid line represents the rectangular region containing the focal
		point, and the dashed line represents the learned approximation model.
	}
\end{figure}
\subsection{Previous Work}
We specifically focuses on \emph{local} and \emph{model-agnostic} methods.
In this section,
we briefly review existing research on local model-agnostic explanations,
particularly focusing on studies closely related to our proposed method.
\subsubsection[LIME]{%
	LIME (Local Interpretable Model-agnostic Explanations)
	\cite{ribeiro2016why}
}
LIME locally
approximates a trained black-box classifier $f: \mathbb{R}^m \to \{0,1\}$
around a focal point $x \in \mathbb{R}^m$
by a linear classifier $g: \mathbb{R}^m \to \{0,1\}$
(\cref{fig:lime}).
The process involves:
\begin{enumerate}
	\item Obtaining a set of perturbed samples $\mathcal{Z}_p$ around $x$
	      and the set of pseudo-labels $f(\mathcal{Z}_p) = \{f(z) \mid z \in \mathcal{Z}_p\}$.
	\item Learning a linear classifier $g$
	      using $\mathcal{Z}_p$ and $f(\mathcal{Z}_p)$
	      by minimizing the following loss function:
	      \begin{equation}
		      \label{eq:lime_loss}
		      \mathcal{L}(f,g,\pi_x)=\sum_{z\in\mathcal{Z}_p}
		      \pi_x(z){\left(f(z)-g(z)\right)}^2,
	      \end{equation}
	      where $\pi_x(z)$ is a weight function designed to be larger for samples
	      closer to $x$, typically implemented using an exponential kernel.
\end{enumerate}

LIME provides valuable insights into the local behavior of the model
by showing the contribution of each feature to the output $f(x)$.
However, due to not explicitly indicating the perturbation region,
users cannot assess the effective scope of the explanation
\cite{ribeiro2018anchors}.

\subsubsection[Anchor]{%
	Anchor~\cite{ribeiro2018anchors}
}\label{sec:anchor}
Anchor represents a rectangular region containing the focal point $x$,
expressed as a conjunction of feature predicates (a rule),
that maximizes the probability of the black-box classifier $f$
outputting $f(x)$ within the region.
It aims to highlight important features
contributing significantly to the output (\cref{fig:anchor}).

For a discrete $m$-dimensional input space $\ispace$
with a trained black-box classifier $f: \ispace \to \{0,1\}$,
an instance $x \in \ispace$,
and a distribution $\mathcal{D}$ over the input space,
a rule $A(z) = a_{i_1}(z) \wedge a_{i_2}(z) \wedge \dots \wedge a_{i_t}(z)$ is defined.
The predicates $a_i(z)$ evaluate to true ($=1$) when $z_i = x_i$ and false ($=0$) otherwise.
The accuracy $\Prec(A)$ and coverage $\Cov(A)$ of the rule $A$ are defined as follows:
\begin{align}
	\Prec(A) & =\mathbb{E}_{z\sim\mathcal{D}(z|A)}
	[\mathbbm{1}_{f(z)=f(x)}], \label{eq:anchor_def_prec} \\
	\Cov(A)  & =\mathbb{E}_{z\sim\mathcal{D}(z)}[A(z)].
\end{align}

where $\mathcal{D}(z|A)$ is the conditional distribution in the region
where the rule $A$ returns true.
$\Prec(A)$ represents the probability that the output of $f$ matches
between the perturbation $z\sim\mathcal{D}(z|A)$ and the focal point $x$,
and $\Cov(A)$ expresses the probability that the perturbation $z$ fits into $A$.

Anchor maximizes coverage under the constraint that
the accuracy of the rule $A$ exceeds a given threshold $\tau$.
However, \cref{eq:anchor_def_prec} is not directly computable.
Introducing a confidence level $1-\delta$ $(0\le\delta\le1)$,
the accuracy constraint is relaxed as follows:
\begin{equation}
	\label{eq:const_prec}
	P(\Prec(A)\ge\tau)\ge1-\delta.
\end{equation}
Thus, the following optimization problem is solved:
\begin{equation}
	\label{eq:main_problem}
	A^*=\underset{A \text{~s.t. } P(\Prec(A)\ge\tau)\ge1-\delta\wedge A(x)=1}
	{\arg\max}\Cov(A).
\end{equation}

@@

\ifnum\mode=1
	\subsection{Computational Complexity}
	Post-hoc explanation methods including LIME, Anchor, and R-LIME
	need to query the black-box model multiple times
	to get labels for perturbed samples, which is computationally expensive.
	The number of queries required for LIME is $|\mathcal{Z}_p|$,
	where $|\mathcal{Z}_p|$ is the number of perturbed samples.
	On the other hand,
	the number of queries required for Anchor and R-LIME is
	$\mathcal{O}(B\cdot m^2+m^2\cdot\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]})$,
	where $B$ is the beam width, $m$ is the number of features, and
	$\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}$ is the computational complexity of
	pure exploration
	of the multi-armed bandit algorithm with $B\cdot m$ arms and beam width $B$.
	$\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}$ depends on hyperparameters
	such as threshold $\tau$, confidence level $1-\delta$ and tolerance $\epsilon$.
\fi

@@

In contrast, R-LIME expresses the application scope of explanations
as a conjunction of predicates.
For example, the explanation for instance A
under $\tau=0.70$ (\cref{fig:A-rlime-70}) indicates that it is applicable
only to married prisoner (Married$=$Yes).
Furthermore, the generated explanations include the accuracy and coverage of
the approximation region, allowing users to evaluate how reliable the
explanations are.
For example, the coverage of the explanation for instance B under $\tau=0.90$
(\cref{fig:B-rlime-90}) is 0.01\%,
indicating that the decision boundaries around instance B are complex,
making it challenging to obtain a high-accuracy linear approximation.
This allows users to discern that the application scope of this explanation is
very narrow, limiting its utility.

\subsection{Quantitative Evaluation}\label{sec:exp2}
\subsubsection{Experimental Setup}
To demonstrate that R-LIME learns a highly accurate linear approximation model
in the optimized approximation region,
we conducted a comparison of the local accuracy of explanations
between LIME and R-LIME\@.
Using the same settings as in \cref{sec:exp_setting},
we randomly sampled 100 instances from the test data of the recidivism dataset
and generated explanations (with $\tau=0.70,0.80,0.90$) using LIME and R-LIME\@.
We then sampled 10,000 instances within the rectangular region
obtained by R-LIME and calculated the local accuracy of both methods.
\subsubsection{Experimental Results}
\begin{figure}[tbp]
	\centering
	\includegraphics[width=0.55\textwidth]{experiments/exp2/box_plot}
	\caption[Comparison of Local Accuracy between R-LIME and LIME]{%
		Comparison of local accuracy between LIME and R-LIME\@.
		R-LIME achieved higher and less variable accuracy compared to LIME
		for all values of $\tau$.
	}\label{fig:box_plot}
\end{figure}
The results are presented in \cref{fig:box_plot},
showcasing the distribution of the accuracy of the linear approximation models
learned by LIME and R-LIME\@.
R-LIME exhibits higher accuracy compared to LIME for all values of $\tau$.
This suggests that the linear classifiers learned by LIME and R-LIME
differ significantly,
and R-LIME learns a high-accuracy linear classifier adapted to the rectangular region.
Additionally, as $\tau$ increases,
the variability in the accuracy of LIME widens.
This indicates that the linear classifiers learned by LIME may not function
effectively as approximation models depending on how the region is selected.

\section{Challenges and Future Perspectives}
 {%
  \def\imgwidth{0.45\textwidth}
  \begin{figure}[t]
	  \centering
	  \begin{subfigure}[t]{\imgwidth}
		  \centering
		  \includegraphics[width=0.8\textwidth]{visual-rlime3}
		  \caption{R-LIME for balanced label distribution.}\label{fig:balanced}
	  \end{subfigure}
	  \begin{subfigure}[t]{\imgwidth}
		  \centering
		  \includegraphics[width=0.8\textwidth]{visual-rlime-imbalanced}
		  \caption{R-LIME for imbalanced label distribution.
		  }\label{fig:imbalanced}
	  \end{subfigure}
	  \caption[Behavior of R-LIME for balanced and imbalanced label distribution]{%
		  Behavior of R-LIME for balanced and imbalanced label distribution.
		  In case of imbalanced label distribution,
		  the approximation region covers the entire input space and the
		  linear approximation model always outputs the majority label.
	  }
  \end{figure}
 }

>>>>

\section{Proposed Method}
\begin{figure}[tbp]
	\centering
	\begin{subfigure}[t]{0.3\textwidth}
		\centering
		\includegraphics[width=\textwidth]{visual-lime}
		\caption{%
			LIME: Locally approximates the decision boundary around the focal point.
		}\label{fig:lime}
	\end{subfigure}%
	\hspace{0.03\textwidth}
	\begin{subfigure}[t]{0.3\textwidth}
		\centering
		\includegraphics[width=\textwidth]{visual-anchor}
		\caption{%
			Anchor: Maximizes coverage of a rectangular region
			containing the focal point under accuracy constraints.
		}\label{fig:anchor}
	\end{subfigure}
	\hspace{0.03\textwidth}
	\begin{subfigure}[t]{0.3\textwidth}
		\centering
		\includegraphics[width=\textwidth]{visual-rlime3}
		\caption{%
			R-LIME:
			Maximizes coverage of a rectangular region
			under lower constraints on the accuracy of the linear classifier.
		}\label{fig:rlime}
	\end{subfigure}
	\caption[Visual comparison of LIME, Anchor and R-LIME]{%
		Visual comparison of LIME, Anchor and R-LIME (our proposed method).
		The solid line represents the rectangular region containing the focal
		point, and the dashed line represents the learned approximation model.
	}
\end{figure}
\subsection{Previous Work}
We specifically focuses on \emph{local} and \emph{model-agnostic} methods.
In this section,
we briefly review existing research on local model-agnostic explanations,
particularly focusing on studies closely related to our proposed method.
\subsubsection[LIME]{%
	LIME (Local Interpretable Model-agnostic Explanations)
	\cite{ribeiro2016why}
}
LIME locally
approximates a trained black-box classifier $f: \mathbb{R}^m \to \{0,1\}$
around a focal point $x \in \mathbb{R}^m$
by a linear classifier $g: \mathbb{R}^m \to \{0,1\}$
(\cref{fig:lime}).
The process involves:
\begin{enumerate}
	\item Obtaining a set of perturbed samples $\mathcal{Z}_p$ around $x$
	      and the set of pseudo-labels $f(\mathcal{Z}_p) = \{f(z) \mid z \in \mathcal{Z}_p\}$.
	\item Learning a linear classifier $g$
	      using $\mathcal{Z}_p$ and $f(\mathcal{Z}_p)$
	      by minimizing the following loss function:
	      \begin{equation}
		      \label{eq:lime_loss}
		      \mathcal{L}(f,g,\pi_x)=\sum_{z\in\mathcal{Z}_p}
		      \pi_x(z){\left(f(z)-g(z)\right)}^2,
	      \end{equation}
	      where $\pi_x(z)$ is a weight function designed to be larger for samples
	      closer to $x$, typically implemented using an exponential kernel.
\end{enumerate}
LIME provides valuable insights into the local behavior of the model
by showing the contribution of each feature to the output $f(x)$.
However, due to not explicitly indicating the perturbation region,
users cannot assess the effective scope of the explanation
\cite{ribeiro2018anchors}.

\subsubsection[Anchor]{%
	Anchor~\cite{ribeiro2018anchors}
}\label{sec:anchor}
Anchor represents a rectangular region containing the focal point $x$,
expressed as a conjunction of feature predicates (a rule),
that maximizes the probability of the black-box classifier $f$
outputting $f(x)$ within the region.
It aims to highlight important features
contributing significantly to the output (\cref{fig:anchor}).
For a discrete $m$-dimensional input space $\ispace$
with a trained black-box classifier $f: \ispace \to \{0,1\}$,
an instance $x \in \ispace$,
and a distribution $\mathcal{D}$ over the input space,
a rule $A(z) = a_{i_1}(z) \wedge a_{i_2}(z) \wedge \dots \wedge a_{i_t}(z)$ is defined.
The predicates $a_i(z)$ evaluate to true ($=1$) when $z_i = x_i$ and false ($=0$) otherwise.
The accuracy $\Prec(A)$ and coverage $\Cov(A)$ of the rule $A$ are defined as follows:
\begin{align}
	\Prec(A) & =\mathbb{E}_{z\sim\mathcal{D}(z|A)}
	[\mathbbm{1}_{f(z)=f(x)}], \label{eq:anchor_def_prec} \\
	\Cov(A)  & =\mathbb{E}_{z\sim\mathcal{D}(z)}[A(z)].
\end{align}
where $\mathcal{D}(z|A)$ is the conditional distribution in the region
where the rule $A$ returns true.
$\Prec(A)$ represents the probability that the output of $f$ matches
between the perturbation $z\sim\mathcal{D}(z|A)$ and the focal point $x$,
and $\Cov(A)$ expresses the probability that the perturbation $z$ fits into $A$.
Anchor maximizes coverage under the constraint that
the accuracy of the rule $A$ exceeds a given threshold $\tau$.
However, \cref{eq:anchor_def_prec} is not directly computable.
Introducing a confidence level $1-\delta$ $(0\le\delta\le1)$,
the accuracy constraint is relaxed as follows:
\begin{equation}
	\label{eq:const_prec}
	P(\Prec(A)\ge\tau)\ge1-\delta.
\end{equation}
Thus, the following optimization problem is solved:
\begin{equation}
	\label{eq:main_problem}
	A^*=\underset{A \text{~s.t. } P(\Prec(A)\ge\tau)\ge1-\delta\wedge A(x)=1}
	{\arg\max}\Cov(A).
\end{equation}

@@

\ifnum\mode=1
	\subsection{Computational Complexity}
	Post-hoc explanation methods including LIME, Anchor, and R-LIME
	need to query the black-box model multiple times
	to get labels for perturbed samples, which is computationally expensive.
	The number of queries required for LIME is $|\mathcal{Z}_p|$,
	where $|\mathcal{Z}_p|$ is the number of perturbed samples.
	On the other hand,
	the number of queries required for Anchor and R-LIME is
	$\mathcal{O}(B\cdot m^2+m^2\cdot\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]})$,
	where $B$ is the beam width, $m$ is the number of features, and
	$\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}$ is the computational complexity of
	pure exploration problem of finding the $B$ best arms among $B\cdot m$ arms.
	$\mathcal{O}_{\mathrm{MAB}[B\cdot m,B]}$ depends on hyperparameters
	such as threshold $\tau$, confidence level $1-\delta$ and tolerance $\epsilon$.
\fi

@@

In contrast, R-LIME expresses the application scope of explanations
as a conjunction of predicates.
For example, the explanation for instance A
under $\tau=0.70$ (\cref{fig:A-rlime-70}) indicates that it is applicable
only to married prisoner (Married$=$Yes).
Furthermore, the explanations include the accuracy and coverage,
allowing users to evaluate its reliability.
For example, the coverage of the explanation for instance B under $\tau=0.90$
(\cref{fig:B-rlime-90}) is 0.01\%,
indicating that the decision boundaries around instance B are complex,
making it challenging to obtain a high-accuracy linear approximation.
This allows users to discern that the application scope of this explanation is
very narrow, limiting its utility.

\subsection{Quantitative Evaluation}\label{sec:exp2}
\subsubsection{Experimental Setup}
To demonstrate that R-LIME learns a highly accurate linear approximation model
in the optimized approximation region,
we conducted a comparison of the local accuracy of explanations
between LIME and R-LIME\@.
Using the same settings as in \cref{sec:exp_setting},
we randomly sampled 100 instances from the test data of the recidivism dataset
and generated explanations (with $\tau=0.70,0.80,0.90$) using LIME and R-LIME\@.
We then sampled 10,000 instances within the rectangular region
obtained by R-LIME and calculated the local accuracy of both methods.
\subsubsection{Experimental Results}
\begin{figure}[tbp]
	\centering
	\includegraphics[width=0.53\textwidth]{experiments/exp2/box_plot}
	\caption[Comparison of Local Accuracy between R-LIME and LIME]{%
		Comparison of local accuracy between LIME and R-LIME\@.
		R-LIME achieved higher and less variable accuracy compared to LIME
		for all values of $\tau$.
	}\label{fig:box_plot}
\end{figure}
The results are presented in \cref{fig:box_plot},
showing the distribution of the accuracy of the linear approximation models
learned by LIME and R-LIME\@.
R-LIME achieved higher accuracy compared to LIME for all values of $\tau$.
This suggests that the linear classifiers learned by LIME and R-LIME
differ significantly,
and R-LIME learns a high-accuracy linear classifier adapted to the rectangular region.
Additionally, as $\tau$ increases,
the variability in the accuracy of LIME widens.
This indicates that the linear classifiers learned by LIME may not function
effectively as approximation models depending on how the region is selected.

\section{Challenges and Future Perspectives}
 {%
  \def\imgwidth{0.47\textwidth}
  \begin{figure}[t]
	  \centering
	  \begin{subfigure}[t]{\imgwidth}
		  \centering
		  \includegraphics[width=0.75\textwidth]{visual-rlime3}
		  \caption{R-LIME for balanced label distribution.}\label{fig:balanced}
	  \end{subfigure}
	  \begin{subfigure}[t]{\imgwidth}
		  \centering
		  \includegraphics[width=0.75\textwidth]{visual-rlime-imbalanced}
		  \caption{R-LIME for imbalanced label distribution.
		  }\label{fig:imbalanced}
	  \end{subfigure}
	  \caption[Behavior of R-LIME for balanced and imbalanced label distribution]{%
		  Behavior of R-LIME for balanced and imbalanced label distribution.
		  In case of imbalanced label distribution,
		  the approximation region covers the entire input space and the
		  linear approximation model always outputs the majority label.
	  }
  \end{figure}
 }
